Foundations of ecological and evolutionary change

### 2025-06-05 foundations_of_ecoevo.Rmd
UPdate two of the references

The interdependence of ecological and evolutionary processes has long been recognised [e.g., @Darwin1859; @Fisher1958; @Pelletier2009], but the rise of eco-evolutionary models, which incorporate both, is relatively recent following a widespread recognition that ecology and evolution can happen on similar timescales [@Govaert2019; @Yamamichi2023].
Currently, a universally recognised formal definition of eco-evolutionary change is lacking, with some theoreticians broadly interpreting "eco-evolutionary dynamics" to allow for a separation of ecological and evolutionary timescales [@Lion2023] and others advocating for a more narrow interpretation in which no such separation is permitted [@Bassar2021].
@Bassar2021 identify two types of eco-evolutionary models that follow from these interpretations.
The first type uses separate equations to model population change versus evolutionary change, thereby allowing for any number of ecological, evolutionary, or environmental feedbacks [e.g., @Lion2018; @Patel2018; @Lion2023].
The second type models population demographics as functions of quantitative traits, with ecological and evolutionary change following from demographic processes and trait distributions operating on the same timescale [e.g., @Barfield2011; @Simmonds2020; @Jaggi2024]. 
Both model types can be very general, but like all predictive models, they rely on simplifying assumptions for tractability [@Levins1966; @Luque2016].
These simplifying assumptions are often grounded in the Price equation to demonstrate accuracy and logical consistency when modelling evolutionary change [e.g., @Coulson2008; @Barfield2011; @Rees2016; @Lion2018].
For example, @Barfield2011 link their model back to @Price1970, which they consider to be a "universal law of evolution", to place their conclusions concerning stage structured evolution in the broader context of evolutionary theory. 
The role of fundamental equations is therefore important for unifying theory [@Luque2016], and we believe that a fundamental equation of eco-evolutionary change is needed.

@@

> **Box 1:** The Price equation is an abstract formula to represent evolutionary change. Formulated originally in the early 1970s by George Price [@Price1970; @Price1972], it postulates some basic properties that all evolutionary systems must satisfy: change over time, ancestor and descendant relations, and a character or phenotype [@Rice2004]. Using simple algebraic language, the Price equation represents evolutionary change with the predominant notation, $$\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).$$ In the above equation, $\Delta\bar{z}$ is the change in the average character value $z$ over a time step of arbitrary length, $w$ is an individual's absolute fitness, and $\bar{w}$ average population fitness. On the right-hand side of the equation, the first term is the covariance between a character value $z$ and fitness $w$, which reflects $\bar{z}$ change attributable to differential survival and reproduction. The second term is the expected value of $\Delta z$, which reflects the extent to which offspring deviate from parents in $z$ [@Rice2004; @Okasha2006; @Frank2012]. A more specific version of the covariance term was already known within the quantitative and population genetics tradition [@Robertson1966], usually representing the action of natural selection. The Price equation adds an expectation term and abstracts away from any specific mechanisms of replication or reproduction, or mechanisms of inheritance. Its definitional nature and lack of substantive biological assumptions has been portrayed both as a strength [@Rice2004; @Frank2012; @Luque2016; @Baravalle2022], and its greatest weakness. The abstract nature of the Price equation places it at the top of the hierarchy of fundamental theorems of evolution from which the rest (Robertson's theorem, Fisher's fundamental theorem, breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be easily derived [@Queller2017; @Lehtonen2016; @Lehtonen2018]. This abstractness is also key to developing a more general view of evolution [@Rice2020; @Luque2021; @Edelaar2023]. In contrast, some researchers consider the Price equation just a triviality (even tautological), and useless without further modelling assumptions [@VanVeelen2005; @VanVeelen2012]. The debate remains open [@VanVeelen2020; @Baravalle2024].

@@

We can now interpret $\Omega$ as the population size at $t+1$, $N_{t+1}$ and note that $N\beta$, $N\delta$, and $N$ are the total births, total deaths, and size of the population at $t$, respectively,

@@

If we assume that individuals are identical, then we can drop the subscript $i$ such that $\beta_{i} = \beta$ and $\delta_{i} = \delta$.
Summing from 1 to $N$, we can rewrite the above,

@@

In some cases, we might also consider $\Omega$ to be the total contribution of a population to ecosystem function.
For example, $z$ might be interpreted as trait that underlies a function of interest such as carbon capture or photosynthetic rate.
When that the relevant characteristic ($z_{i}$) is additive across individuals  (e.g., the total amount of carbon capture in a population is the sum of carbon captured across individuals), $\Omega$ could be an appropriate metric of ecosystem function.
This will not always be the case.
The sum of some measurable traits such as leaf pH [@Cornelissen2011] will not be ecologically meaningful when summed across individuals.
And in other cases, the relationship between $\Omega$ and some ecosystem function of interest is likely to be subadditive.
For example, we could interpret $z_{i}$ as the frequency at which a pollinator $i$ visits plants (which might change from $t$ to $t+1$ for any number of reasons, including ontogenic or environmental, as reflected in $\Delta z_{i}$).
While $\Omega$ would then be the total frequency of pollinator visits in a population, pollinator saturation effects would likely have a logarithmic relationship between $\Omega$ and total seed set [e.g., @Mu2018].
While we have argued that ecological and evolutionary change can both be derived from the same fundamental relationship between birth, death, and individual characteristics, the same is not true for ecosystem function.
Nevertheless, Box 3 provides a toy example of four plants with different fitnesses and carbon captures.
We use this toy example to demonstrate a concrete example of our proposed framework.

>>>>

The interdependence of ecological and evolutionary processes has long been recognised [e.g., @Darwin1859; @Fisher1958; @Pelletier2009], but the rise of eco-evolutionary models, which incorporate both, is relatively recent following a widespread recognition that ecology and evolution can happen on similar timescales [@Govaert2019; @Yamamichi2023].
Currently, a universally recognised formal definition of eco-evolutionary change is lacking, with some theoreticians broadly interpreting "eco-evolutionary dynamics" to allow for a separation of ecological and evolutionary timescales [@Lion2023] and others advocating for a more narrow interpretation in which no such separation is permitted [@Bassar2021].
@Bassar2021 identify two types of eco-evolutionary models that follow from these interpretations.
The first type uses separate equations to model population change versus evolutionary change, thereby allowing for any number of ecological, evolutionary, or environmental feedbacks [e.g., @Lion2018; @Patel2018; @Lion2023].
The second type models population demographics as functions of quantitative traits, with ecological and evolutionary change following from demographic processes and trait distributions operating on the same timescale [e.g., @Barfield2011; @Simmonds2020; @Jaggi2024]. 
Both model types can be very general, but like all predictive models, they rely on simplifying assumptions for tractability [@Levins1966; @Luque2016].
These simplifying assumptions are often grounded in the Price equation to demonstrate accuracy and logical consistency when modelling evolutionary change [e.g., @Coulson2008; @Barfield2011; @Rees2016; @Lion2018].
For example, @Barfield2011 link their model back to @Price1970, which they consider to be a "universal law of evolution", to place their conclusions concerning stage structured evolution in the broader context of evolutionary theory. 
The role of fundamental equations is therefore important for unifying theory [@Luque2021], and we believe that a fundamental equation of eco-evolutionary change is needed.

@@

> **Box 1:** The Price equation is an abstract formula to represent evolutionary change. Formulated originally in the early 1970s by George Price [@Price1970; @Price1972], it postulates some basic properties that all evolutionary systems must satisfy: change over time, ancestor and descendant relations, and a character or phenotype [@Rice2004]. Using simple algebraic language, the Price equation represents evolutionary change with the predominant notation, $$\bar{w}\Delta\bar{z} = \mathrm{Cov}\left(w, z\right) + \mathrm{E}\left(w\Delta z\right).$$ In the above equation, $\Delta\bar{z}$ is the change in the average character value $z$ over a time step of arbitrary length, $w$ is an individual's fitness, and $\bar{w}$ average population fitness. On the right-hand side of the equation, the first term is the covariance between a character value $z$ and fitness $w$, which reflects $\bar{z}$ change attributable to differential survival and reproduction. The second term is the expected value of $\Delta z$, which reflects the extent to which offspring deviate from parents in $z$ [@Rice2004; @Okasha2006; @Frank2012]. A more specific version of the covariance term was already known within the quantitative and population genetics tradition [@Robertson1966], usually representing the action of natural selection. The Price equation adds an expectation term and abstracts away from any specific mechanisms of replication or reproduction, or mechanisms of inheritance. Its definitional nature and lack of substantive biological assumptions has been portrayed both as a strength [@Rice2004; @Frank2012; @Luque2016; @Baravalle2022], and its greatest weakness. The abstract nature of the Price equation places it at the top of the hierarchy of fundamental theorems of evolution from which the rest (Robertson's theorem, Fisher's fundamental theorem, breeder's equation, Hamilton's rule, adaptive dynamics, etc.) can be easily derived [@Queller2017; @Lehtonen2016; @Lehtonen2018]. This abstractness is also key to developing a more general view of evolution [@Rice2020; @Luque2021; @Edelaar2023]. In contrast, some researchers consider the Price equation just a triviality (even tautological), and useless without further modelling assumptions [@VanVeelen2005; @VanVeelen2012]. The debate remains open [@VanVeelen2020; @Baravalle2024].

@@

We can now interpret $\Omega$ as the population size at $t+1$, $N_{t+1}$,

@@

If we assume that individuals are identical, then we can drop the subscript $i$ such that $\beta_{i} = \beta$ and $\delta_{i} = \delta$ and note that $N\beta$, $N\delta$, and $N$ are the total births, total deaths, and size of the population at $t$, respectively,

Summing from 1 to $N$, we can rewrite the above,

@@

In some cases, we might also consider $\Omega$ to be the total contribution of a population to ecosystem function.
For example, $z$ might be interpreted as a trait that underlies an ecosystem function of interest such as carbon capture or photosynthetic rate.
When that the relevant characteristic ($z_{i}$) is additive across individuals  (e.g., the total amount of carbon capture in a population is the sum of carbon captured across individuals), $\Omega$ could be an appropriate metric of ecosystem function.
This will not always be the case.
Some measurable traits such as leaf pH [@Cornelissen2011] will not be ecologically meaningful when summed across individuals.
And in other cases, the relationship between $\Omega$ and some ecosystem function of interest is likely to be subadditive.
For example, we could interpret $z_{i}$ as the frequency at which a pollinator $i$ visits plants (which might change from $t$ to $t+1$ for any number of reasons, including ontogenic or environmental, as reflected in $\Delta z_{i}$).
While $\Omega$ would then be the total frequency of pollinator visits in a population, pollinator saturation effects would likely have a logarithmic relationship between $\Omega$ and total seed set [e.g., @Mu2018].
Hence, while we have argued that ecological and evolutionary change can both be derived from the same fundamental relationship between birth, death, and individual characteristics, the same is not true for ecosystem function.
Nevertheless, Box 3 provides a toy example of four plants with different fitnesses and carbon captures.
We use this toy example to demonstrate a concrete example of our proposed framework.
